\bf Estimating the number of clusters using Cross Validation

### 2016-04-08 New simulation/technometrics.tex
Brain tumor data: add preprocessing detail and discussion

The third data set is gene expression data of $5$ types of brain tumours from \cite{de2008clustering}, which contains $42$ observations including $10$ medulloblastomas, $10$ malignant gliomas, $10$ atypical teratoid/rhabdoid tumours (AT/RTs), $8$ primitive neuroectodermal tumours (PNETs) and $4$ normal cerebella. It was originally used by \cite{pomeroy2002prediction} to show that medulloblastomas are molecularly distinct from other types of brain tumours. There were $7129$ genes originally, but after feature selection, $1379$ genes left. So the data set is $42 \times 1379$, which serves as a high dimensional example.

\begin{table}[H]
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table2} Number of clusters selected by each algorithm}
\begin{tabular}{lccccc}
    \hline                  
 & Congress Voting && Breast Cancer && Brain Tumours \\ \hline                    
CH-index & $2$ && $2$  && $2$    \\
Hartigan & $3$ && $3$  &&  $4$   \\
Jump & $10$ && $9$ && $1$  \\   
Prediction strength & $2$ &&  $2$ && $1$    \\
Bootstrap stability & $2$ &&  $2$ &&  $7$   \\ 
Gap & $8$ && $10$ && $10$  \\   
Gaussian-Mix & $2$ && $5$ && $2$  \\   
Gabriel & $2$  && $3$  && $5$  \\    
Wold & $2$ && $3$ && $4$ \\  \hline

@@

Since most congressmen vote base on their parties' interest, $2$ parties
(Democratic and Republican) represent two clusters in this data set. So the
optimal number should be two. Close inspection shows $k$-means with
$k=2$ separates the two parties very well with the lowest miss-classification
error $10.43\%$ ($k=3$ has $14.78\%$). Note that CH-index and Bootstrap
stability also return $k=2$, but $2$ is the lower bound those methods can
select for $k$. So it's not clear they actually choose $k=2$ or they hit the
lower bound (they would pick $k=1$ if allowed). For the breast cancer data,
it's known to have at least $2$ cluster based on whether it's benign specimens
or not. But it doesn't mean the optimal $k$ should be $2$.
\cite{fujita2014non} noticed that the malign group is quite heterogeneous and
can be further clustered into at least two subgroups. Hence, the result $k=3$
given by our proposed Gabriel method make sense. For
the brain tumours gene expression data set which is high-dimension, it is also clear that 
the correct number of clusters should be $5$, each cluster corresponding to one type of 
brain tumour. From the table we can see only our proposed
Gabriel method correctly selects $k=5$, underline how
difficult it was to pick the right $k$ when dimension is high.

>>>>

The third data set is gene expression data of $5$ types of brain tumours from \cite{de2008clustering}, which contains $42$ observations including $10$ medulloblastomas, $10$ malignant gliomas, $10$ atypical teratoid/rhabdoid tumours (AT/RTs), $8$ primitive neuroectodermal tumours (PNETs) and $4$ normal cerebella. It was originally used by \cite{pomeroy2002prediction} to show that medulloblastomas are molecularly distinct from other types of brain tumours.

Data was collected with Affymetrix microarrays, which estimates the number of RNA copies found in the cell sample (frozen specimens). Data was preprocessed as following: as a common procedure, all genes with expression level below $10$ are set to a minimum level threshold of $10$. The maximum threshold is set at $16,000$. Values below or above these thresholds are often unreliable. To remove uninformative genes, for each gene $j$ (attribute), the mean $m_j$ among the samples was calculated. The $10\%$ largest and smallest values are discarded in order to remove extreme values. Based on the mean, following transformation is applied on every value $x_{ij}$ of sample $i$ and gene j
\[	y_{ij} = \log_2 (x_{ij}/m_j)	\] 
Genes with expression levels differing by at least $l$-flod in at least $c$ samples from their mean expression level across the sample were selected. The parameter $l$ and $c$ were chosen so as to produce a filtered data set with at least $10\%$ of the original number of genes \citep{de2008clustering}. Such preprocessing procedure results in $1379$ genes selected. After filtering step, data is restored in the original scale (i.e. $x_{ij}$). So the data set is $42 \times 1379$, which serves as a high dimensional example.

\begin{table}[H]
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table2} Number of clusters selected by each algorithm}
\begin{tabular}{lccccc}
    \hline                  
 & Congress Voting && Breast Cancer && Brain Tumours \\ \hline                    
CH-index & $2$ && $2$  && $2$    \\
Hartigan & $3$ && $3$  &&  $4$   \\
Jump & $10$ && $9$ && $1$  \\   
Prediction strength & $2$ &&  $2$ && $1$    \\
Bootstrap stability & $2$ &&  $2$ &&  $7$   \\ 
Gap & $8$ && $10$ && $10$  \\   
Gaussian-BIC & $2$ && $5$ && $2$  \\   
Gabriel & $2$  && $3$  && $5$  \\    
Gabriel-corr-correct & $2$  && $2$  && $5$  \\  
Wold & $2$ && $3$ && $4$ \\  \hline

@@

Since most congressmen vote based on their parties' interest, two parties
(Democratic and Republican) represent two clusters in this data set. So the
optimal number should be two. Close inspection shows $k$-means with
$k=2$ separates the two parties very well with the lowest miss-classification
error $10.43\%$ ($k=3$ has $14.78\%$). Our proposed Gabriel CV method, along with 
several other algorithms correctly pick $k=2$ for this data set. Note that CH-index and Bootstrap
stability also return $k=2$, but $2$ is the lower bound those methods can
select for $k$. So it's not clear they actually choose $k=2$ or they hit the
lower bound (they would pick $k=1$ if allowed).

For the breast cancer data, it's known to have $2$ clusters based on 
whether it's benign specimens or malignant specimens. Although the proposed Gabriel method
picks $k=3$, its correlation-corrected version correctly picks $k=2$. It shows the necessity 
to correct for the high correlation among data features. However, for this data $k=3$ is also
an acceptable answer, as \cite{fujita2014non} noticed that the malign group is quite heterogeneous and
can be further clustered into at least two subgroups.

The brain tumours gene expression data is a typical high-dimension data with $p > n$. It is clear that 
the correct number of clusters should be $5$, the number of brain tumour types. Indeed, $k$-means results with $k=5$ show that the $4$ normal human cerebella are well separated from brain tumours as they form a single cluster.  
Medulloblastomas, malignant gliomas and AT/RT tumours also form $3$ respective clusters, which means they can be separated from each other. Such results are similar to those found in \cite{pomeroy2002prediction}, whose analysis
is based on the first $3$ principle components of the original data.    
From table \ref{table2} we can see only our proposed Gabriel method (as well as its correlation-corrected version) correctly selects $k=5$, underline how difficult it was to pick the right $k$ when dimension is high.
